\bf Model Visualization Techniques for a Social Network Model

### 2017-08-12 PaperToDo.Rmd
senate data example

@@

- [ ] Write an abstract
- [ ] Submit to journal
- [ ] Spacing around figures WAY to large above/below captions

@@

- [ ] Add histogram of number of tie changes from $t_1$ to $t_2$ in simulations 
- [ ] Add another example of a data set for M-in-DS (Sect 4.1) (GoT?)
- [ ] Make gif available online
- [ ] Insert link to full gif in paper
- [X] Figure 6: reverse order of adjacency matrix y-axis
- [X] Figure 7 not showing up??
- [X] Fix error on p. 21 (probably related to above)
- [X] Heatmap p 23 missing caption
- [ ] Describe Figure 8 more. What is going on??
- [ ] Add geomnet example in 4.2
- [X] p. 26-28 figure refs & code is broken
- [X] Get rid of repeated plot p 29
- [X] p. 34 code is broken
- [ ] General proof reading

>>>>

Don't spend any time placing the figures!!!!! Place figure just in front of the paragraph of text describing it.

@@

- [ ] Write an abstract - Gelman method/sanity check. First sentence of every paragraph of the intro & the very last sentence of intro. 
- [ ] Submit to journal
- [ ] Spacing around figures WAY to large above/below captions - change fig.height for coord_fixed figures 
- [ ] Make axes labels and text larger in figures
- [ ] Make labels of points larger

@@

- [ ] Add histogram of number of tie changes from $t_1$ to $t_2$ in simulations 
- [ ] Add another example of a data set for M-in-DS (Sect 4.1) (Congress)
- [ ] Make gif available online
- [ ] Insert link to full gif in paper
- [X] Figure 6: reverse order of adjacency matrix y-axis
- [ ] Figure 5: get rid of Vs on node labels, make font bigger
- [X] Figure 7 not showing up??
- [X] Fix error on p. 21 (probably related to above)
- [X] Heatmap p 23 missing caption
- [ ] Describe Figure 8 more. What is going on??
- [ ] Add geomnet example in 4.2
- [X] p. 26-28 figure refs & code is broken
- [X] Get rid of repeated plot p 29
- [X] p. 34 code is broken
- [ ] General proof reading

### 2017-08-12 paper.Rnw
senate data example

<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.8\\linewidth', par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation - LETS GO/SAOM-removing-blindfold/")
@

@@

The \texttt{RSiena} software that fits the models to data also comes equipped with the \texttt{sienaGOF()} function for examining model fit. This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}. Examples of the auxiliary statistics are the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine. The fit is tested using Mahalanobis distance to determine a $p$-value when comparing the statistics calculated from the observed data (excluding the first observation that is conditioned on) to the distribution of the statistics when calculated on the simulated waves. So, good fit is acheived when $p$ is large. XX more detail about mahalanobis distance and calculating p-value XX

The \texttt{plot.sienaGOF()} function allows us to visualize this fit. Below is an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model. Box plots and violin plots are drawn on top of each other in Figure~\ref{fig:sienagof} in order to show the distribution of the simulated network observations compared to the true data shown in red points connected by red lines. If the red points lie well within the simulated values, the model is a good fit to the data. The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.

<<sienagof, fig.cap="Goodness of fit measures for two network statistics: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data.", cache = TRUE, message = FALSE, warning = FALSE>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship")
gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship")
p1 <- plot(gof1)
p2 <- plot(gof2)
gridExtra::grid.arrange(p1, p2)
@

@@

To guide our visual exploration of stochastic actor-oriented models, we used a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage. These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy, with changes that are more noticeable. The subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves. This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network adjacency matrices in Figure~\ref{fig:adjmatviz}. For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves. We will also be working with one actor level categorical covariate, drinking behavior. This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week. The actor covariates and networks are visualized in Figures~\ref{fig:adjmatviz} and~\ref{fig:alldataviz}. 
<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data. The subset we will be using is outlined in red.", fig.width=8, results='hide',fig.show='asis'>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
ggplot(data= view_rs2, aes(x = x, y=y)) + 
  geom_tile(aes(fill = as.factor(value))) + 
  scale_fill_manual("i likes j", labels=c("no", "yes"), 
                    values = c('white', 'black')) +
  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
            aes(xmin = 19.5, xmax = 35.5, ymin = 19.5, ymax = 35.5, fill =NA)) + 
  facet_wrap(~wave) + theme_bw() +
  labs(x = "i", y = "j") + 
  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank()) 
@

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", fig.width=8>>=
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1 
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat, 
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 3, vjust = .5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
  theme_net() +
  facet_wrap(~Wave, labeller = "label_both") + 
  theme(legend.position = 'bottom', panel.background = element_rect(fill = NA, color = "grey30"))
alldataviz
@

>>>>

<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.8\\linewidth', par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
@

@@

The \texttt{RSiena} software that fits the models to data also comes equipped with the \texttt{sienaGOF()} function for examining model fit. This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}. Examples of the auxiliary statistics are the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine. The fit is tested using Mahalanobis distance to determine a $p$-value when comparing the statistics calculated from the observed data (excluding the first observation that is conditioned on) to the distribution of the statistics when calculated on the simulated waves. So, good fit is acheived when $p$ is large. \st{XX more detail about mahalanobis distance and calculating p-value XX}

The \texttt{plot.sienaGOF()} function allows us to visualize this fit. Below is an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model. Box plots and violin plots are drawn on top of each other in Figure~\ref{fig:sienagof} in order to show the distribution of the simulated network observations compared to the true data shown in red points connected by red lines. If the red points lie "well within" the simulated values, the model is a good fit to the data. The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.

<<sienagof, fig.cap="Goodness of fit measures for two network statistics: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data. \st{The p-value shown on the x-axis is .... }", cache = TRUE, message = FALSE, warning = FALSE>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship")
gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship")
p1 <- plot(gof1)
p2 <- plot(gof2)
gridExtra::grid.arrange(p1, p2)
@

@@

To guide our visual exploration of stochastic actor-oriented models, we used a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage. These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy, with changes that are more noticeable. The subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves. This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network adjacency matrices in Figure~\ref{fig:adjmatviz}. For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves. We will also be working with one actor level categorical covariate, drinking behavior. This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week. The actor covariates and networks are visualized in Figures~\ref{fig:adjmatviz} and~\ref{fig:alldataviz}. 
<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data. The subset we will be using is outlined in red.", fig.width=8, results='hide',fig.show='asis', fig.width=8, fig.height=3>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
ggplot(data= view_rs2, aes(x = x, y=y)) + 
  geom_tile(aes(fill = as.factor(value))) + 
  scale_fill_manual("i likes j", labels=c("no", "yes"), 
                    values = c('white', 'black')) +
  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
            aes(xmin = 19.5, xmax = 35.5, ymin = 19.5, ymax = 35.5, fill =NA)) + 
  facet_wrap(~wave) + theme_bw() +
  labs(x = "i", y = "j") + 
  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank()) 
@

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", fig.width=8, fig.height=3>>=
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1 
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat, 
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 3, vjust = .5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
  theme_net() +
  facet_wrap(~Wave, labeller = "label_both") + 
  theme(legend.position = 'bottom', panel.background = element_rect(fill = NA, color = "grey30"))
alldataviz
@
